---
title: "Dynamical System Components"
format: html
---

```{r}
library(ramp.xds)
```

In developing a modular framework for model building, we identified five core processes. 
In `ramp.xds`, these are handled in three chunks, called ***COMPONENTS.*** 

-   **XH** -- Human / Vertebrate Hosts (see [The **XH** Component](XHcomponent.qmd))

    -   $\cal X$ -- a state space for infection dynamics, immunity, disease, and related processes;

    -   $\cal H$ -- a set of processes to represent host demographic processes, including births, deaths and aging

-   **MY** -- Adult mosquitoes (see [The **MY** Component](MZ))

    -   $\cal Y$ -- a state space describing infection dynamics

    -   $\cal M$ -- a state space describing mosquito ecology

-   **L** -- Aquatic mosquitoes (see [The **L** Component](Lcomponent.qmd))

    -   $\cal L$ -- a state space describing aquatic mosquito ecology

The first step in setting up a model is to determine what module to use for each component. 

Each chunk handles its own tasks, such as setting up and indexing variables, computing the derivatives, and parsing outputs. 
Each *COMPONENT* has its own design elements to handle a specific set of challenges associated with model building to serve the needs of malaria analysts. 
Some of these functions are also useful for research on malaria and other mosquito-transmitted pathogens. 


### **XH** Component

The **X** Component is specified by setting `Xname =` *`model_name`*, where *`model_name`* is a string that defines the `class` for S3 methods.

For example:

```{r}
sis_mod <- xds_setup(Xname = "SIS")
```

In basic setup, the **H** component is set to a trivial model with no births and no deaths. 

### **MY** Component

The **MY** Component is specified by setting `MYZname =` *`model_name`*, where *`model_name`* is a string that defines the `class` for S3 methods, or equivalently, a model family in **`ramp.xds`** or **`ramp.library`**

For example:

```{r}
macdonald_mod <- xds_setup(MYZname = "macdonald")
```

### **L** Component

The **L** Component is specified by setting `Lname =` *`model_name`*, where *`model_name`* is a string that defines the `class` for S3 methods.

For example:

```{r}
aqua_mod <- xds_setup(Lname = "basicL")
```
