---
title: "Install SimBA"
subtitle: "Getting Started: Vignette 1"
editor:
  mode: source
bibliography: ../refs/refs.bib
csl: ../refs/plos.csl
html-math-method:
  method: mathjax
  url: "https://cdn.jsdelivr.net/npm/mathjax@3/es5/tex-mml-chtml.js"
---

**SimBA** is a suite of [R packages](Rpackages.qmd) on github. At some point, we will make them available on CRAN, but for now you will need to install them directly from github using an R package called [devtools](https://cran.r-project.org/web/packages/devtools/index.html).

> NOTE: When you install R packages, you might have to install some additional packages that they depend on. This can take some time, so please be patient.

# devtools

This code will install and load devtools.

```{r}
#| eval: false
install.packages("devtools")
library(devtools)
```

# ramp.xds

Core computation for SimBA is done by ramp.xds.

If you’ve installed and loaded `devtools`. These commands download and install `ramp.xds`.

```{r}
#| eval: false
devtools::install_github("dd-harp/ramp.xds")
library(ramp.xds)
```

# The Rest

This will install (but not load) the other packages:

```{r}
#| eval: false
devtools::install_github("dd-harp/ramp.library")
devtools::install_github("dd-harp/ramp.func")
devtools::install_github("dd-harp/ramp.forcing")
devtools::install_github("dd-harp/ramp.demog")
devtools::install_github("dd-harp/ramp.work")
devtools::install_github("dd-harp/ramp.qa")
```

# Next

After installing the software, we recommend the next vignette: [Build a Model](demo_build.qmd)
