---
title: "Build a Model"
subtitle: "Getting Started: Vignette 2"
editor:
  mode: source
bibliography: ../refs/refs.bib
csl: ../refs/plos.csl
html-math-method:
  method: mathjax
  url: "https://cdn.jsdelivr.net/npm/mathjax@3/es5/tex-mml-chtml.js"
---

This vignette assumes:

you have learned about mathematical models for malaria;

you know how to use R and RStudio;

you have [installed](demo_install.qmd) **`ramp.xds`** and **`ramp.library`**.

# Basic Setup

SimBA has a basic utility called xds_setup that handles the first stage of setup – we call it basic setup – for models for malaria epidemiology, transmission dynamics, and control that were formulated as dynamical systems. The software is capable of supporting systems of differential equations or discrete time systems.

To build this model, we need two packages:

```{r}
library(ramp.xds)
library(ramp.library)
```

## `xds_setup()`

These two software packages have modules that implement published models for malaria and mosquito ecology.

For the purposes of this demonstration, we will start with a compartmental model for malaria called the SIP model. For mosquitoes, we use the SI model.

Setting up a model is as easy as calling xds_setup using the names of the modules:

```{r}
model <- xds_setup(Xname = "SIP", MYname = "SI")
```

If you want to know more about xds_setup you can ask for help:

```{r}
#| eval: false
?xds_setup
```

You can also read more about it in [Basic Setup](basic_setup.qmd).

## The xds Model Object

The object that was returned by xds_setup, here called model, is called an xds model object. In R, it is a list:

```{r}
typeof(model) 
```

In fact, is a structured list with a defined set of elements, many of which are also lists. The class of the xds object is xds_obj

```{r}
#| eval: false
class(model) 
```

When you type the name of an object, the default behavior for R is to call print, which dispatches on the class of the object. print.xds_obj is defined in ramp.xds.It returns a description of the model and its structure:

```{r}
model
```

Notice that the model has a trivial **L** module. How did that happen? `xds_setup` has defined default. In this case, `xds_setup` used the default option because we did not tell it what to do.

The `xds` object has is set up to handle a lot of complexity, including structures that will not get used unless we choose to do it. It’s there in case we want to add a feature to the model. To see how it is constructed, we can examine the names of objects attached to model:

```{r}
names(model) 
```

From the user’s perspective, it won’t be necessary to know most of this. Feel free to poke around, but it’s probably easier to use built-in functions to change parameters, get outputs, change initial values, etc. We suggest tucking the information away in case there come a time when it might be useful.

# Next

After learning how to build a model, we recommend:

- The `ramp.xds` vignette [Getting Started](https://dd-harp.github.io/ramp.xds/articles/GettingStarted.html); or

- the next vignette in the Basics series: [Solve a Model](demo_solve.qmd); or

- take a deeper dive into [Basic Setup](basic_setup.qmd) to figure out how to set up a model with the features you want.
