---
title: "Demo"
format: revealjs
bibliography: "./data/refs.bib"
csl: "./data/plos.csl" 
---


## Welcome 

This assumes familiarity with R, dynamical systems & malaria. 

To get started, install  **`ramp.xds`** from the GitHub repository using `devtools::install_github()`: 

```{r, echo=T, eval=F}
library(devtools)
devtools::install_github("dd-harp/ramp.xds")
``` 

Then load it:

```{r, echo=T}
library(ramp.xds)
```

## A Model 

In **`ramp.xds`**, it's easy to build, solve, and visualize malaria models, formulated as dynamical systems:  

```{r, fig.height=4, fig.width=12, echo=T}
demo_mod <- xds_setup()
demo_mod <- xds_solve(demo_mod)
par(mfrow = c(1,2))
xds_plot_EIR(demo_mod)
xds_plot_PR(demo_mod)
```
## The **`xds`** Object  

The object that was returned by `xds_setup()`, that we named `demo_mod`, is a compound list called an **`xds`** object. Three text strings define the `class` of each **dynamic module**: 

```{r, echo=T}
demo_mod$Xname
```

```{r, echo=T}
demo_mod$MYZname
```

```{r, echo=T}
demo_mod$Lname
```

## Setup:: Defaults 

`xds_setup()` supplies default values: 

+ default dynamic modules (*i.e.* `SIS`, `macdonald`, and `trivial`) recreate a model published in 1982 [@AronJL1982PopulationDynamics]; 

+ each dynamic module sets up its own default parameters & initial values; 

+ all structural parameters are set to $1$ (*i.e.*, the number of patches, aquatic habitats, host strata, host species, & vector species)

In a few slides, we'll describe how to change the defaults.  

## The **`xds`** Object : **X** module 

The **X** component -- a dynamical module describing **infection dynamics & immunity in humans** or other vertebrate hosts -- is dispatched by `Xpar[[i]]`. Basic setup sets up only the first host species:

```{r, echo=T}
demo_mod$Xpar[[1]]
```

## The **`xds`** Object : **MYZ** module 

The **MYZ** component defines **adult mosquito ecology and infection dynamics** for each vector species, stored in `MYZpar[[i]]`: 

```{r, echo=T}
class(demo_mod$MYZpar[[1]])
```

## The **`xds`** Object : **L** module 

The **L** component defines **aquatic mosquito population dynamics**  for each vector species, 
stored in `Lpar[[i]]`: 

```{r, echo=T}
class(demo_mod$Lpar[[1]])
```

The `trivial` module is a function that emergence rate of adult mosquitoes, $\Lambda(t).$ Here, the scaling parameter is called `Lambda` 

```{r, echo=T}
demo_mod$Lpar[[1]]$Lambda
```

'## Trivial modules & Forcing

To implement plug-and-play modularity, each component has a `trivial` module. All `trivial` modules return a function of time:

$$s \times F_S(t) \times F_T(t)$$

+ $s$ is a scale parameter named in context (*e.g.* `Lambda`)

+ $F_S(t)$ is the seasonal pattern, returned by `F_season` 

+ $F_T(t)$ is an inter-annual trend, returned by `F_trend`  

Some function families can be set up by [`make_function`](https://dd-harp.github.io/ramp.xds/reference/make_function.html)

## Setup :: `xds_setup`  

Setting up basic features is easy using `xds_setup_*()`  

+ [`xds_setup()`](https://dd-harp.github.io/ramp.xds/reference/xds_setup.html) can be used for most situations 

+ [`xds_setup_cohort()`](https://dd-harp.github.io/ramp.xds/reference/xds_setup_cohort.html) sets up a  human / vertebrate host model forced by  $F_{E}(a,t),$ 
a function that returns the daily EIR as a function of age and time. 

+ Other `xds_setup_*()` functions streamline special cases and define a `frame` for solving, described in the  [documentation](https://dd-harp.github.io/ramp.xds/reference/index.html#build-setup) 

## Setup :: Dynamic Modules 

+ Dynamic modules are configured by passing: 

    + `Xname = "model_name"` 
    + `MYZname = "model_name"` 
    + `Lname = "model_name"` 

+ The values in a named list overwrite default parameters & initial values: 
    + `Xopts = list(X=1, r=1/180)`  
    + `MYZopts = list(M=1, g=1/5)`
    + `Lopts = list(L=1, psi = 1/5)` 


## Setup :: Structural Elements 

`xds_setup()` sets up a model with one host and one vector species: 

+ `nPatches` is the number of spatial units for adult mosquitoes

+ `membership` indexes the patch membership of each habitat, passed as a vector where $\mbox{max}($`membership`$)\leq$`nPatches` and `nHabitats`$=\mbox{length}($`membership`$)$

+ `HPop` is the population size of the strata, 
`residence` indexes the patches where the humans reside, and  
`nStrata`$=\mbox{length}($`HPop`$)=
\mbox{length}($`residence`$).$ 

## Setup :: Spatial Dynamics 

## Solving 

It's also easy to visualize the outputs:

```{r, fig.height=4, fig.width=12, echo=T}
par(mfrow = c(1,2))
xds_plot_EIR(demo_mod)
xds_plot_PR(demo_mod)
```

## References