---
title: "Module Library"
bibliography: ../refs/refs.bib
csl: ../refs/plos.csl
subtitle: "A Catalogue of Modules"
format: html
---

------------------------------------------------------------------------

*To run a module in SimBA, it must be added to the **model library.** In SimBA, a **model family** that has been encoded and added to the model library is called a **module.***

------------------------------------------------------------------------

The model catalogue for SimBA

## **X** Modules

A list of the currently available **XH** modules and their locations can be accessed inline: 

```{r}
#| eval: false
?XH_module_list
```

These modules are classified as being either [**malaria** modules](malaria_epidemiology.qmd) or not. Modules for [**other pathogens**](other_X_modules.qmd). 
 

## **MY** -- Adult Mosquitoes

A list of the currently available **MY** modules and their locations can be accessed inline: 

```{r}
#| eval: false
?MY_module_list
```

## **L** -- Aquatic Mosquitoes

A list of the currently available **L** modules and their locations can be accessed inline: 

```{r}
#| eval: false
?L_module_list
```

