---
title: "Mosquito Dispersal"
editor:
  mode: source
author: 
  - name:  "David L Smith"  
    url: https://faculty.washington.edu/smitdave/
    orcid: "0000-0003-4367-3849" 
    email: "smitdave@uw.edu"
    affiliations: "University of Washington"
bibliography: ../refs/refs.bib
csl: ../refs/plos.csl
---

In **`ramp.xds`**, mosquito dispersal can be configured as part of [**basic setup**](basic_setup.qmd), or it can be changed later using `setup_K_matrix`, which handles setup for models with minimal ecology as well as behavioral state models. 

This vignette covers:

+ [Dispersal Kernels](@sec-kernel) --- the standard implementation of dispersal kernels across all **MY** modules

+ [`setup_K_matrix()`](@sec-setup) --- a discussion of `setup_K_matrix()` and its methods:

    - `matrix` --- set up $K$ by passing an $N_p \times N_p$ matrix with the proper form
    
    - `herethere` --- disperse evenly to all other patches 
    
    - `xy` --- use the $x,y$ locations and a spatial kernel
    
    - `zero` --- set the matrix to all zeros
    
    - `N` --- for behavioral state models, set up multiple kernels

+ [`xds_setup()`](@sec-basic) --- how `xds_setup()` handles dispersal kernels

+ [Examples](@sec-examples)

Inline help is available:
```{r}
#| eval: false
?xds_info_mosquito_dispersal
?setup_K_matrix
```

# Dispersal Kernels {#sec-kernel}

Let $k_{i,j}$ describe the fraction of mosquitoes emigrating from each patch that land on another patch: ${k}_{i,j} (t) \in K(t)$ is the fraction of mosquitoes leaving patch $j$ that end up in patch $i$. Each column in the matrix describes mosquitoes leaving *from* $j.$ 

Emigration rates and population *loss* associated with dispersal are handled separately (see [Mosquito Demography](mosquito_demography.qmd)), so by convention, each column sums up to zero. 

\begin{equation}
K = 
\left[
\begin{array}{ccccc}
j=1&j=2&j=3&\cdots&j=p \\ 
\boxed{
\begin{array}{c}
-1 \\
{k}_{2,1} \\
{k}_{3,1} \\
\vdots \\
{k}_{p,1} \\
\end{array}}  
&
\boxed{
\begin{array}{c}
{k}_{1,2} \\
-1 \\
{k}_{3,2} \\
\vdots \\
{k}_{p,2} \\
\end{array}} 
&
\boxed{
\begin{array}{c}
{k}_{1,3} \\
{k}_{3,2} \\
-1 \\
\vdots \\
{k}_{p,3} \\
\end{array}} 
&
\boxed{
\begin{array}{c}
\cdots \\
\cdots \\
\cdots \\
\ddots \\
\cdots 
\end{array}}
&
\boxed{
\begin{array}{c}
{k}_{1,p} \\
{k}_{2,p} \\
{k}_{3,p} \\
\vdots \\
-1 \\
\end{array}} 
\end{array}
\right]
\end{equation}


Behavioral state models often have two or more dispersal matrices. 




# `setup_K_matrix` {#sec-setup}

```{r}
library(ramp.xds)
```

The function is called as:

```{r}
#| eval: false 
setup_K_matrix(name, xds_obj, options=list(), s=1)
``` 

The function `setup_K_matrix` dispatches on `options` but before function is dispatched, a block of code is run that makes it possible to pass arguments in several ways. 

but `class(name)` is checked and changed before the function call: 

+ The function was designed to be called by passing a string called `name`, the name of the S3 method: 
  - if `name` is a string, then `class(options) <- 'name'`
  - `options` is a named list with arguments for `setup_K_matrix.name`
+ Matrix 
    - if `name` is a matrix, then `class(options) <- 'as_matrix'`
    - it is equivalent to calling `setup_K_matrix("as_matrix", options = )
+ 
+ if `name` is a list with an item called `name`  then `class(options) <- 'name$name'`

+ a string: 


## `as_matrix`

```{r}
M = matrix(c(-1, .97, .03, 
             .95, -1, .05, 
             0.02, .98, -1), 3, 3)
M
```

```{r}
colSums(M)< 1e-10
```

If the matrix is passed to `setup_K` then (after passing checks) it becomes the time spent matrix. 

```{r}
model <- xds_setup(nPatches = 3)
model <- setup_K_matrix(M, model)
get_K_matrix(model)
```

## herethere

If `name = "herethere"` then the fraction moving to every other patch is equal.  
```{r}
model <- setup_K_matrix("herethere", model)
get_K_matrix(model)
```

## xy 

The `setup_timspent.xy` option grabs the residence vector and then calls `make_timespent_xy` with options: 

+ `xy` --- the $x$ and $y$ coordinates of each patch

+ `kern` --- a function to weight patches by distance, any information on the `xds` model object (*e.g.* human population density), or any options passed to the function. The form of the function is:
$$F_k(d, M, O)$$ where `M` is the model object and `O` is `options` is a named list. 

+ k_opts --- options, passed as a named list (to `setup_timepent(name = "xy", ...)`).

In the following example, we generate 10 random  patch locations, as $x,y$ coordinates, in a $10 \times 10$ box centered at 0. The human population has 10 strata: one for each patch.  Individuals spend, on average, around 80% of their time at home (drawn from a `beta` distribution). The remaining time is allocated to points by a distance function: $$F_K(d) = e^{-(d-3)^2}.$$

## Behavioral State Models
 
The dispersal kernels for **behavioral state models** are the result of searching for different resources. For *basic ecology* models, the dispersal kernel is called `K_matrix`, but *behavioral state models* can have two or more kernels, each with different names. 

To handle this, the functions `get_K_matrix` and `change_K_matrix` have an extra argument `which_K`. By default, for basic ecology, the default is set to "K" and it can be ignored. It can not be ignored when working with behavioral state models. 

The behavioral state models are in `ramp.library` so we load it. 

```{r}
library(ramp.library)
```

The `BQ` module is a behavioral state module. We'll set one up to work with: 

```{r}
modBQ <- xds_setup(MYname = "BQ", nPatches=3)
```

To look at the $K_b$ matrix (for blood searching), we use `get_K_matrix` with the second argument, `Kb`

```{r}
get_K_matrix(modBQ, "Kb")
```
We can call `setup_K_matrix` to change it, but we must add `which_K = "Kb"` to the options list: 

```{r}
modBQ <- setup_K_matrix("herethere", modBQ, options=list(which_K = "Kb"))
get_K_matrix(modBQ, "Kb")
```
A method for `setup_K_matrix` was written for behavioral state models. The method name is "N", and the argument must say now many kernels it will be setting up: $N=2$. The options must also have an argument called `opts` which is a list of length $N$: each one describes how to set up the kernel. For example:  

```{r}
Kopts = list(name = "N", N=2, opts = list()) 
Kopts$opts[[1]] = list(name = "herethere", which_K = "Kb")
Kopts$opts[[2]] = list(name = "herethere", which_K = "Kq")
```

Now, we can set up a model and then modify both kernels at the same time. After basic setup, the kernels are both set to $0$ matrices. 

```{r}
modBQ4 <- xds_setup(MYname = "BQ", nPatches=4)
get_K_matrix(modBQ4, "Kq")
```
Now, we pass the options, and we get: 

```{r}
modBQ4 <- setup_K_matrix("N", modBQ4, Kopts)
get_K_matrix(modBQ4, "Kq")
```

```{r}
get_K_matrix(modBQ4, "Kb")
```
Note that we could also pass `Kopts` to `xds_setup`

```{r}
modBQ5 <- xds_setup(MYname = "BQ", nPatches=5, Koptions=Kopts)
get_K_matrix(modBQ5, "Kb")
```
# Basic Setup {#sec-basic}

The time spent matrix can be configured in basic setup.  


## Defaults

The default setting for `xds_setup` is that mosquitoes don't move. If $N_p=1$ then the time spent matrix for one patch is:

```{r}

model <- xds_setup()
get_K_matrix(model)
```

For multiple strata: 

```{r}
model <- xds_setup(nPatches=3)
get_K_matrix(model)
```

At setup, a parameter list called `Koptions` can be passed that calls `setup_K_matrix`.  Any list of options that would set up a kernel using `setup_K_matrix` will work here. For example: 

```{r}
model <- xds_setup(nPatches=3, Koptions = list(name="herethere"))
get_K_matrix(model)
```

# Examples {#sec-examples}

## A Spatial Kernel

The example is meant to be illustrative, not realistic. 


```{r, fig.height=8, fig.width=8}
#| eval: false
set.seed(2302)
x = runif(10, -5, 5)
y = runif(10, -5, 5)
xy = data.frame(x=x, y=y)
kern = function(d, model, options){exp(-(d-3)^2)}
options = list(name="xy", xy=xy, ker=kern)
model <- xds_setup(nPatches=10, HPop = sample(2^c(1:10)), 
                   residence = 1:10,
                   membership = 1:10,
                   Loptions = list(Lambda = sample(2^c(1:10))), 
                   Koptions=options)
#model <- setup_K_matrix("xy", model, options)
model <- xds_solve(model)
model <- last_to_inits(model)
model <- xds_analyze(model)
model <- make_VC(model)
VC <- get_VC_matrix(model)

plot_connectivity_graph(VC, xy, mn=0.01, clrs = viridisLite::turbo(10),
                        arr.pos=0.8, scl=.1)
```