---
title: "Mosquito Dispersal Kernels"
editor: source
author: 
  - name:  "David L Smith"  
    url: https://faculty.washington.edu/smitdave/
    orcid: "0000-0003-4367-3849" 
    email: "smitdave@uw.edu"
    affiliations: "University of Washington"
bibliography: ../refs/refs.bib
csl: ../refs/plos.csl
---

*** 

*This vignette discusses how to set up a matrix describing mosquito dispersal. The mosquito dispersal matrix. *

***

Mosquito dispersal is understood as the outcome of *flight bouts* and *searching* for resources [@ServiceMW1997LongShort;@WuSL2020MBITES;@SanchezCHM2026MosquitoDispersal]. In SimBA's patch-based framework, mosquito dispersal has two parts: emigration rates and dispersal kernels. 

The mathematical framework was developed to model emigration rates in relation to the *local availability* of resources [@WuSL2023SpatialDynamics] --- mosquitoes will tend to stay in patches where all the resources are readily available [@SanchezCHM2026MosquitoDispersal]. Emigration and the hazards associated with dispersal are probably an important source of mosquito mortality [@SaulA2003ZooprophylaxisZoopotentiation]. Knowing that a mosquito has *left* a patch is one thing, but the models must also say where they land, which can be affected by specific features on a landscape [@BidlingmayerWL1974MosquitoFlight;@BidlingmayerWL1979MosquitoDipteral;@BidlingmayerWL1981MosquitoFlight;@BidlingmayerWL1985MeasurementAdult;@LutambiAM2013MathematicalModelling]. If mosquito dispersal is the result of a *flight bouts,* where the direction and distances traveled are affected by wind [@BidlingmayerWL1985PreliminaryStudy;@MidegaJT2012WindDirection;@PerkinsTA2013HeterogeneityMixing;@WuSL2020MBITES;@SanchezCHM2026MosquitoDispersal], then a mosquito leaving a patch could end up anywhere, not just in a neighboring patch. This calls for a general framework to model dispersal among patches. The framework thus handles mosquito dispersal in two parts: *emigration* and *emigration loss* are one part; and the destinations of emigrating mosquitoes that survive is the other. 

This vignette describes **mosquito dispersal kernels** --- a matrix describing where the surviving emigrants end up.  Emigration rates and population *loss* associated with dispersal are handled separately (see [Mosquito Demography](mosquito_demography.qmd)), so by convention, each column sums up to zero. 

***

In **`ramp.xds`**, a mosquito dispersal kernel can be configured as part of [**basic setup**](basic_setup.qmd), or it can be changed later using `setup_K_matrix`, which handles setup for models with minimal ecology as well as behavioral state models. 

This vignette covers:

+ [Dispersal Kernels](@sec-kernel) --- the standard implementation of dispersal kernels across all **MY** modules

+ [`setup_K_matrix()`](@sec-setup) --- a discussion of `setup_K_matrix()` and its methods:

    - `matrix` --- set up $K$ by passing an $N_p \times N_p$ matrix with the proper form
    
    - `herethere` --- disperse evenly to all other patches 
    
    - `xy` --- use the $x,y$ locations and a spatial kernel
    
    - `zero` --- set the matrix to all zeros
    
    - `N` --- for behavioral state models, set up multiple kernels

+ [`xds_setup()`](@sec-basic) --- how `xds_setup()` handles dispersal kernels

+ [Examples](@sec-examples)

# Emigration & Loss

We let $\sigma$ denote the patch emigration rate. The parameter $\mu$ describes population loss that is conditioned on emigration from a patch: a fraction $\mu$ never reaches another patch.  If $g$ is the mortality rate (in each patch), then mortality and migration terms are summarized by a single matrix, $\Omega,$ called the [mosquito demographic matrix](mosquito_demography.qmd)

$$\Omega = \mbox{diag}\left(g + \sigma \mu\right) - K \cdot \mbox{diag}\left(\sigma \left(1-\mu\right) \right)$$
The dispersal kernel $K$ is called the *dispersal kernel* or the `K_matrix.` For more information, see [Mosquito Demography](mosquito_demography.qmd). 

# Dispersal Kernels {#sec-kernel} 

Let $k_{i,j}$ describe the fraction of mosquitoes emigrating from each patch that land on another patch: ${k}_{i,j} (t) \in K(t)$ is the fraction of mosquitoes leaving patch $j$ that end up in patch $i$. Each column in the matrix describes mosquitoes leaving *from* $j.$ The columns of $K$ sum to 0. 

\begin{equation}
K = 
\left[
\begin{array}{ccccc}
j=1&j=2&j=3&\cdots&j=p \\ 
\boxed{
\begin{array}{c}
-1 \\
{k}_{2,1} \\
{k}_{3,1} \\
\vdots \\
{k}_{p,1} \\
\end{array}}  
&
\boxed{
\begin{array}{c}
{k}_{1,2} \\
-1 \\
{k}_{3,2} \\
\vdots \\
{k}_{p,2} \\
\end{array}} 
&
\boxed{
\begin{array}{c}
{k}_{1,3} \\
{k}_{3,2} \\
-1 \\
\vdots \\
{k}_{p,3} \\
\end{array}} 
&
\boxed{
\begin{array}{c}
\cdots \\
\cdots \\
\cdots \\
\ddots \\
\cdots 
\end{array}}
&
\boxed{
\begin{array}{c}
{k}_{1,p} \\
{k}_{2,p} \\
{k}_{3,p} \\
\vdots \\
-1 \\
\end{array}} 
\end{array}
\right]
\end{equation}


Behavioral state models often have two or more dispersal matrices. 


## `setup_K_matrix` {#sec-setup}

```{r}
library(ramp.xds)
```

The function used to configure the dispersal kernels is called `setup_K_matrix.` The function call is:

```{r}
#| eval: false
setup_K_matrix(name, xds_obj, options=list(), s=1)
``` 

Inline help is available: 
```{r}
#| eval: false
?xds_info_mosquito_dispersal
?setup_K_matrix
``` 

The function `setup_K_matrix` dispatches on `options` but before function is dispatched, a block of code is run that makes it possible to pass arguments in several ways. 

but `class(name)` is checked and changed before the function call: 

+ The function was designed to be called by passing a string called `name`, the name of the S3 method: 
  - if `name` is a string, then `class(options) <- 'name'`
  - `options` is a named list with arguments for `setup_K_matrix.name`
+ Matrix 
    - if `name` is a matrix, then `class(options) <- 'as_matrix'`
    - it is equivalent to calling `setup_K_matrix("as_matrix", options = )

+ if `name` is a list with an item called `name`  the function dispatches on `setup_K_matrix.list`. 



### `as_matrix`

The `as_matrix` option makes it possible to build a matrix using any method and use it as a `K_matrix`, so long as it has the right form. The matrix is checked by `check_K_matrix`

```{r}
#| eval: false
?check_K_matrix
```


```{r}
my_M = matrix(c(-1, .97, .03, 
             .95, -1, .05, 
             0.02, .98, -1), 3, 3)
my_M
```

```{r}
check_K_matrix(my_M, 3)
```

After setting up a model, the matrix can be set up in two ways. 

```{r}
model <- xds_setup(nPatches = 3)
```

The call to `setup_K_matrix` using the standard syntax is:

```{r}
model1 <- setup_K_matrix("as_matrix", model, list(K_matrix=my_M))
get_K_matrix(model1)
```
An easier way is to pass the matrix instead of the name: 

```{r}
model2 <- setup_K_matrix(my_M, model)
get_K_matrix(model2)
```

### `herethere`

If `name = "herethere"` then the fraction moving to every other patch is equal. There are no parameters: 

```{r}
model <- setup_K_matrix("herethere", model)
get_K_matrix(model)
```

### `xy` 

The `setup_timspent.xy` option grabs the residence vector and then calls `make_K_matrix_xy` with options: 

+ `xy` --- the $x$ and $y$ coordinates of each patch

+ `F_K` --- a function to weight relative travel among patches by distance with the form
$$F_K(d, V)$$
+ `V` --- a set of variables to pass to $F_K$ 


For example, we generate 10 random points: 

```{r}
#| echo: false 
set.seed(2302)
```

```{r}
nXY=20
model_xy = xds_setup(nPatches=nXY)
```

```{r}
xy = data.frame(x=runif(nXY, -5, 5), y=runif(nXY, -5, 5))
```

We construct a function:
$$F_K(d,V) = e^{-(d-3)^2}.$$

```{r}
FdV = function(d, V=list()){exp(-(d-3)^2)}
```

```{r}
K_xy <- make_K_matrix_xy(xy, FdV)
```

```{r, fig.height=5, fig.width=5}
library(ramp.qa)
plot_kernel_graph(K_xy, xy, mn=0.01, clrs = viridisLite::turbo(10),
                        arr.pos=0.7, scl=2)
```

This sets up the model: 

```{r}
#| eval: false
model_xy = setup_K_matrix("xy", model_xy, 
                          list(xy=xy, F_K=FdV, V=list()))

```


### Behavioral State Models
 
The dispersal kernels for **behavioral state models** are the result of searching for different resources. For *basic ecology* models, the dispersal kernel is called `K_matrix`, but *behavioral state models* can have two or more kernels, each with different names. 

To handle this, the functions `get_K_matrix` and `change_K_matrix` have an extra argument `which_K`. By default, for basic ecology, the default is set to "K" and it can be ignored. It can not be ignored when working with behavioral state models. 

The behavioral state models are in `ramp.library` so we load it. 

```{r}
library(ramp.library)
```

The `BQ` module is a behavioral state module. We'll set one up to work with: 

```{r}
modBQ <- xds_setup(MYname = "BQ", nPatches=3)
```

To look at the $K_b$ matrix (for blood searching), we use `get_K_matrix` with the second argument, `Kb`

```{r}
get_K_matrix(modBQ, "Kb")
```
We can call `setup_K_matrix` to change it, but we must add `which_K = "Kb"` to the options list: 

```{r}
modBQ <- setup_K_matrix("herethere", modBQ, options=list(which_K = "Kb"))
get_K_matrix(modBQ, "Kb")
```
A method for `setup_K_matrix` was written for behavioral state models. The method name is "N", and the argument must say now many kernels it will be setting up: $N=2$. The options must also have an argument called `opts` which is a list of length $N$: each one describes how to set up the kernel. For example:  

```{r}
Kopts = list(name = "N", N=2, opts = list()) 
Kopts$opts[[1]] = list(name = "herethere", which_K = "Kb")
Kopts$opts[[2]] = list(name = "herethere", which_K = "Kq")
```

Now, we can set up a model and then modify both kernels at the same time. After basic setup, the kernels are both set to $0$ matrices. 

```{r}
modBQ4 <- xds_setup(MYname = "BQ", nPatches=4)
get_K_matrix(modBQ4, "Kq")
```
Now, we pass the options, and we get: 

```{r}
modBQ4 <- setup_K_matrix("N", modBQ4, Kopts)
get_K_matrix(modBQ4, "Kq")
```

```{r}
get_K_matrix(modBQ4, "Kb")
```

## Basic Setup {#sec-basic}

The time spent matrix can be configured in basic setup.  


### Defaults

The default setting for `xds_setup` is that mosquitoes don't move. If $N_p=1$ then the time spent matrix for one patch is:

```{r}

model <- xds_setup()
get_K_matrix(model)
```

For multiple strata: 

```{r}
model <- xds_setup(nPatches=3)
get_K_matrix(model)
```

### `Koptions` 

At setup, a parameter list called `Koptions` can be passed that calls `setup_K_matrix`.  Any list of options that would set up a kernel using `setup_K_matrix` will work here. For example: 

```{r}
model <- xds_setup(nPatches=3, Koptions = list(name="herethere"))
get_K_matrix(model)
```

Note that for behavioral state models, we could also pass `Kopts` to `xds_setup`

```{r}
modBQ5 <- xds_setup(MYname = "BQ", nPatches=5, Koptions=Kopts)
get_K_matrix(modBQ5, "Kb")
```



# Examples {#sec-examples}

## A Spatial Kernel
