---
title: "The `ramp.xds` Pipeline"
bibliography: ../refs/refs.bib
csl: ../refs/plos.csl
subtitle: "An interactive tutorial: setup → solve → visualize"
format: html
server: shiny
---

------------------------------------------------------------------------

*This tutorial walks through the three-step **`ramp.xds`** pipeline interactively. Choose modules, adjust parameters, and watch the model update in real time.*

------------------------------------------------------------------------

```{r}
#| context: setup
#| include: false
library(ramp.xds)
has_lib <- requireNamespace("ramp.library", quietly = TRUE)
if (has_lib) library(ramp.library)

# --- Module choices (ramp.library modules added when available) ---------------
xh_choices <- c("SIS")
if (has_lib) xh_choices <- c(xh_choices, "SEIS", "SEISd")
xh_choices <- c(xh_choices, "trivial")

my_choices <- c("SI", "macdonald", "GeRM")
if (has_lib) my_choices <- c(my_choices, "SEI", "BQ", "BRQS")
my_choices <- c(my_choices, "trivial")

l_choices <- c("trivial", "basicL")
if (has_lib) l_choices <- c(l_choices, "stages")
```

## Overview

Every **`ramp.xds`** workflow follows three steps:

1.  **Setup** — `xds_setup()` configures a model: choose dynamical modules and set parameters.
2.  **Solve** — `xds_solve()` integrates the system of differential equations.
3.  **Visualize** — built-in plotting functions display the results.

---

## The XH Component — Human Infection Dynamics

Choose a module for the **XH** component, which describes infection dynamics and immunity in the human (or vertebrate host) population.

```{r}
#| echo: false
fluidRow(
  column(4,
    selectInput("Xname", "Xname:", choices = xh_choices, selected = "SIS")
  ),
  column(8, uiOutput("xh_params"))
)
```

---

## The MY Component — Adult Mosquito Ecology & Infection

Choose a module for the **MY** component, which describes adult mosquito ecology and infection dynamics.

```{r}
#| echo: false
fluidRow(
  column(4,
    selectInput("MYname", "MYname:", choices = my_choices, selected = "macdonald")
  ),
  column(8, uiOutput("my_params"))
)
```

---

## The L Component — Aquatic Mosquito Ecology

Choose a module for the **L** component, which describes aquatic-stage mosquito ecology.

```{r}
#| echo: false
fluidRow(
  column(4,
    selectInput("Lname", "Lname:", choices = l_choices, selected = "trivial")
  ),
  column(8, uiOutput("l_params"))
)
```

---

## Solve

Configure the human population and simulation duration, then solve.

```{r}
#| echo: false
fluidRow(
  column(4,
    sliderInput("HPop", HTML("<em>H</em> — human population size:"),
                min = 100, max = 5000, value = 1000, step = 100)
  ),
  column(4,
    sliderInput("Tmax",
                HTML("<em>T</em><sub>max</sub> — simulation duration (days):"),
                min = 60, max = 3650, value = 365, step = 30)
  )
)
```

**Generated code:**

```{r}
#| echo: false
verbatimTextOutput("setup_code")
```

---

## Visualize

```{r}
#| echo: false
tabsetPanel(id = "plots",
  tabPanel("Prevalence (PR)",
    br(), plotOutput("pr_plot", height = "320px"),
    tags$p(tags$small(HTML(
      "Parasite rate (<em>I/H</em>) over time &mdash;
       <code>xds_plot_PR()</code>.")))
  ),
  tabPanel("EIR",
    br(), plotOutput("eir_plot", height = "320px"),
    tags$p(tags$small(HTML(
      "Entomological Inoculation Rate (infectious bites per person per year)
       &mdash; <code>xds_plot_EIR()</code>.")))
  ),
  tabPanel("Mosquito Density (M)",
    br(), plotOutput("M_plot", height = "320px"),
    tags$p(tags$small(HTML(
      "Total adult mosquito density &mdash;
       <code>xds_plot_M()</code>.")))
  ),
  tabPanel("Infected Mosquitoes (Y)",
    br(), plotOutput("Y_plot", height = "320px"),
    tags$p(tags$small(HTML(
      "Infected mosquitoes (pre-EIP) &mdash;
       <code>xds_plot_Y()</code>.")))
  ),
  tabPanel("Infective Mosquitoes (Z)",
    br(), plotOutput("Z_plot", height = "320px"),
    tags$p(tags$small(HTML(
      "Infective mosquitoes (sporozoite-positive) &mdash;
       <code>xds_plot_Z()</code>.")))
  )
)
```

---

```{r}
#| context: server

# =============================================================================
# Dynamic parameter UIs
# =============================================================================

output$xh_params <- renderUI({
  xname <- input$Xname
  switch(xname,
    "SIS" = tagList(
      tags$h5(tags$strong("XH"), tags$code(" — SIS")),
      sliderInput("xh_b", HTML("<em>b</em> — prob. infection per bite:"),
                  min = 0.1, max = 0.9, value = 0.55, step = 0.05),
      sliderInput("xh_r_inv", HTML("1/<em>r</em> — infection duration (days):"),
                  min = 30, max = 365, value = 180, step = 10),
      sliderInput("xh_c", HTML("<em>c</em> — infectiousness to mosquitoes:"),
                  min = 0.05, max = 0.50, value = 0.15, step = 0.05)
    ),
    "SEIS" = tagList(
      tags$h5(tags$strong("XH"), tags$code(" — SEIS")),
      sliderInput("xh_b", HTML("<em>b</em> — prob. infection per bite:"),
                  min = 0.1, max = 0.9, value = 0.55, step = 0.05),
      sliderInput("xh_r_inv", HTML("1/<em>r</em> — infection duration (days):"),
                  min = 30, max = 365, value = 180, step = 10),
      sliderInput("xh_nu_inv", HTML("1/<em>&nu;</em> — latent period (days):"),
                  min = 5, max = 30, value = 14, step = 1),
      sliderInput("xh_c", HTML("<em>c</em> — infectiousness to mosquitoes:"),
                  min = 0.05, max = 0.50, value = 0.15, step = 0.05)
    ),
    "SEISd" = tagList(
      tags$h5(tags$strong("XH"), tags$code(" — SEISd")),
      sliderInput("xh_b", HTML("<em>b</em> — prob. infection per bite:"),
                  min = 0.1, max = 0.9, value = 0.55, step = 0.05),
      sliderInput("xh_r_inv", HTML("1/<em>r</em> — infection duration (days):"),
                  min = 30, max = 365, value = 180, step = 10),
      sliderInput("xh_nu_d", HTML("<em>&nu;</em> — intrinsic incubation (days):"),
                  min = 5, max = 30, value = 14, step = 1),
      sliderInput("xh_c", HTML("<em>c</em> — infectiousness to mosquitoes:"),
                  min = 0.05, max = 0.50, value = 0.15, step = 0.05)
    ),
    "trivial" = tagList(
      tags$h5(tags$strong("XH"), tags$code(" — trivial")),
      sliderInput("xh_kappa", HTML("<em>&kappa;</em> — net infectiousness:"),
                  min = 0.01, max = 0.5, value = 0.1, step = 0.01)
    )
  )
})

output$my_params <- renderUI({
  myname <- input$MYname
  if (myname == "trivial") {
    return(tagList(
      tags$h5(tags$strong("MY"), tags$code(" — trivial")),
      sliderInput("my_Z", HTML("<em>Z</em> — infective mosquitoes:"),
                  min = 0.1, max = 20, value = 1, step = 0.1),
      sliderInput("my_f_triv", HTML("<em>f</em> — blood feeding rate:"),
                  min = 0.1, max = 1.0, value = 0.3, step = 0.05)
    ))
  }
  # Shared parameters for all non-trivial MY modules
  f_default <- if (myname %in% c("BQ", "BQm", "BRQS")) 0.5 else 0.3
  common <- tagList(
    tags$h5(tags$strong("MY"), tags$code(paste0(" — ", myname))),
    sliderInput("my_f", HTML("<em>f</em> — blood feeding rate (per day):"),
                min = 0.1, max = 1.0, value = f_default, step = 0.05),
    sliderInput("my_g_inv", HTML("1/<em>g</em> — mosquito lifespan (days):"),
                min = 3, max = 30, value = 12, step = 1),
    sliderInput("my_eip", HTML("EIP — extrinsic incubation period (days):"),
                min = 5, max = 25, value = 12, step = 1)
  )
  # Extra parameters for BQ-family models
  extra <- NULL
  if (myname %in% c("BQ", "BQm", "BRQS")) {
    extra <- tagList(
      sliderInput("my_sigma_b",
                  HTML("<em>&sigma;<sub>b</sub></em> — emigration (blood-feeding):"),
                  min = 0, max = 0.5, value = 1/8, step = 0.01),
      sliderInput("my_sigma_q",
                  HTML("<em>&sigma;<sub>q</sub></em> — emigration (egg-laying):"),
                  min = 0, max = 0.5, value = 1/8, step = 0.01)
    )
  }
  tagList(common, extra)
})

output$l_params <- renderUI({
  lname <- input$Lname
  switch(lname,
    "trivial" = tagList(
      tags$h5(tags$strong("L"), tags$code(" — trivial")),
      sliderInput("l_Lambda", HTML("&Lambda; — emergence rate (per day):"),
                  min = 100, max = 5000, value = 1000, step = 100)
    ),
    "basicL" = tagList(
      tags$h5(tags$strong("L"), tags$code(" — basicL")),
      sliderInput("l_psi", HTML("<em>&psi;</em> — maturation rate:"),
                  min = 0.01, max = 0.5, value = 1/8, step = 0.01),
      sliderInput("l_phi", HTML("<em>&phi;</em> — mortality rate:"),
                  min = 0.01, max = 0.5, value = 1/8, step = 0.01),
      sliderInput("l_theta", HTML("<em>&theta;</em> — density-dependent mortality:"),
                  min = 0.001, max = 0.1, value = 1/100, step = 0.001)
    ),
    "stages" = tagList(
      tags$h5(tags$strong("L"), tags$code(" — stages")),
      sliderInput("l_psi_E", HTML("<em>&psi;<sub>E</sub></em> — egg hatching rate:"),
                  min = 0.1, max = 1.0, value = 0.5, step = 0.05),
      sliderInput("l_psi1", HTML("<em>&psi;<sub>1</sub></em> — L1 maturation rate:"),
                  min = 0.1, max = 1.0, value = 0.5, step = 0.05),
      sliderInput("l_phi1", HTML("<em>&phi;<sub>1</sub></em> — L1 mortality:"),
                  min = 0.1, max = 1.0, value = 0.5, step = 0.05),
      sliderInput("l_psi_P", HTML("<em>&psi;<sub>P</sub></em> — pupal emergence rate:"),
                  min = 0.1, max = 1.0, value = 1/3, step = 0.05)
    )
  )
})


# =============================================================================
# Build options lists from current inputs
# =============================================================================

xh_opts <- reactive({
  xname <- input$Xname
  switch(xname,
    "SIS" = {
      req(input$xh_b, input$xh_r_inv, input$xh_c)
      list(b = input$xh_b, r = 1/input$xh_r_inv, c = input$xh_c)
    },
    "SEIS" = {
      req(input$xh_b, input$xh_r_inv, input$xh_nu_inv, input$xh_c)
      list(b = input$xh_b, r = 1/input$xh_r_inv,
           nu = 1/input$xh_nu_inv, c = input$xh_c)
    },
    "SEISd" = {
      req(input$xh_b, input$xh_r_inv, input$xh_nu_d, input$xh_c)
      list(b = input$xh_b, r = 1/input$xh_r_inv,
           nu = input$xh_nu_d, c = input$xh_c)
    },
    "trivial" = {
      req(input$xh_kappa)
      list(kappa = input$xh_kappa)
    }
  )
})

my_opts <- reactive({
  myname <- input$MYname
  if (myname == "trivial") {
    req(input$my_Z, input$my_f_triv)
    return(list(Z = input$my_Z, f = input$my_f_triv))
  }
  req(input$my_f, input$my_g_inv, input$my_eip)
  opts <- list(f = input$my_f, g = 1/input$my_g_inv, eip = input$my_eip)
  if (myname %in% c("BQ", "BQm", "BRQS")) {
    req(input$my_sigma_b, input$my_sigma_q)
    opts$sigma_b <- input$my_sigma_b
    opts$sigma_q <- input$my_sigma_q
  }
  opts
})

l_opts <- reactive({
  lname <- input$Lname
  switch(lname,
    "trivial" = {
      req(input$l_Lambda)
      list(Lambda = input$l_Lambda)
    },
    "basicL" = {
      req(input$l_psi, input$l_phi, input$l_theta)
      list(psi = input$l_psi, phi = input$l_phi, theta = input$l_theta)
    },
    "stages" = {
      req(input$l_psi_E, input$l_psi1, input$l_phi1, input$l_psi_P)
      list(psi_E = input$l_psi_E, psi1 = input$l_psi1,
           phi1 = input$l_phi1, psi_P = input$l_psi_P)
    }
  )
})


# =============================================================================
# Build and solve the model
# =============================================================================

mod <- reactive({
  req(input$Xname, input$MYname, input$Lname, input$HPop)
  xds_setup(
    Xname     = input$Xname,
    MYname    = input$MYname,
    Lname     = input$Lname,
    HPop      = input$HPop,
    XHoptions = xh_opts(),
    MYoptions = my_opts(),
    Loptions  = l_opts()
  )
})

solved_mod <- reactive({
  req(input$Tmax)
  xds_solve(mod(), Tmax = input$Tmax, dt = 1)
})


# =============================================================================
# Generated code
# =============================================================================

output$setup_code <- renderText({
  req(input$Xname, input$MYname, input$Lname, input$HPop)
  xo <- xh_opts(); mo <- my_opts(); lo <- l_opts()
  fmt <- function(lst) {
    items <- mapply(function(n, v) paste0(n, " = ", signif(v, 4)),
                    names(lst), lst, USE.NAMES = FALSE)
    paste(items, collapse = ", ")
  }
  paste0(
    'model <- xds_setup(\n',
    '  Xname     = "', input$Xname, '",\n',
    '  MYname    = "', input$MYname, '",\n',
    '  Lname     = "', input$Lname, '",\n',
    '  HPop      = ', input$HPop, ',\n',
    '  XHoptions = list(', fmt(xo), '),\n',
    '  MYoptions = list(', fmt(mo), '),\n',
    '  Loptions  = list(', fmt(lo), ')\n',
    ')\n',
    'model <- xds_solve(model, Tmax = ', input$Tmax, ')'
  )
})


# =============================================================================
# Plots — wrapped in tryCatch for module combinations that lack certain outputs
# =============================================================================

safe_plot <- function(expr) {
  tryCatch(expr, error = function(e) {
    plot.new()
    text(0.5, 0.5, paste("Not available:\n", e$message),
         cex = 1.1, col = "grey50")
  })
}

output$pr_plot  <- renderPlot({ safe_plot(xds_plot_PR(solved_mod(), y01 = TRUE)) })
output$eir_plot <- renderPlot({ safe_plot(xds_plot_EIR(solved_mod())) })
output$M_plot   <- renderPlot({ safe_plot(xds_plot_M(solved_mod())) })
output$Y_plot   <- renderPlot({ safe_plot(xds_plot_Y(solved_mod())) })
output$Z_plot   <- renderPlot({ safe_plot(xds_plot_Z(solved_mod())) })
```