---
title: "Scaling Relationships"
subtitle: "Studying Malaria Metrics Across the Spectrum of Transmission Intensity"
bibliography: data/refs.bib
csl: data/plos.csl 
---

***

Also see:

+ [**ramp.work**:: Scaling for Malaria Metrics](https://dd-harp.github.io/ramp.work/articles/Scaling.html) -- the algorithms explained here are documented in a vignette on the software website. 

+ [**Malaria Theory:: Scaling**](https://faculty.washington.edu/smitdave/theory/scaling.html) -- the algorithms explained here were used to illustrate basic scaling relationships for malaria theory.

*** 

Load the required packages:

```{r}
library(ramp.xds)
library(ramp.library)
library(ramp.work)
library(deSolve)
library(rootSolve)
library(graphics)
```

## xde_scaling

The function `xde_scaling()` defines the relationship between the EIR and the PR, and it outputs stable orbits for each value of aEIR in a mesh from $10^{-1}$ up to $10^{3}$

```{r sis, eval=F}
xds_setup_cohort(Xname = "SIS") -> sis
```

```{r scaling, eval=F}
xde_scaling_eir(sis, 25) -> sis
```

```{r sis scaling, echo=F}
if(file.exists("sis.rda")){
  load("sis.rda")
} else {
  sis <- xds_setup_cohort(Xname = "SIS")
  xde_scaling_eir(sis, 25) -> sis 
  save(sis, file="sis.rda")
}
```

```{r plot, fig.height=3.5, fig.width=5}
plot_eirpr(sis)
```
