---
title: "Stratification"
editor:
  mode: source
bibliography: ../refs/refs.bib
csl: ../refs/plos.csl
format: html
html-math-method:
  method: mathjax
  url: "https://cdn.jsdelivr.net/npm/mathjax@3/es5/tex-mml-chtml.js"
---

In **SimBA,** *stratification* almost always describes a process of segmenting (or subdividing) a population. 

Simple models of malaria assume that everyone is alike, but populations can be heterogeneous in various ways: 

+ heterogeneous exposure and transmission

    - location
    
    - housing quality

+ heterogeneous behaviors

    - care seeking
    
    - drug taking or adherance

+ age and immunity  

+ heterogeneity created by malaria control

    - vaccination

Basic setup asks the user to develop a basic 

Stratification is handled by `ramp.demog`


# Residency

An important aspect of human population strata for most models is the location of home. In spatial models, we make a distinction between where things happen and where things get counted. Information about malaria in humans is often aggregated by the location of home. 


For an ordered list of population strata, the residence vector is a list of patch indices for those strata:
$$\left[i, j, k, ..., y, z \right]$$
The residence matrix, $J,$ is a $N_p \times N_h$ matrix where the $i,j^{th}$ entry is 1 if the $i^{th}$ stratum lives in the $j^{th}$ patch. 



For example, if there are four strata, two live in patch 1 and two in patch 2, then the residence vector is: 
$$\left[1, 1, 2, 2 \right]$$
and the residence matrix is: 

```{r}
library(ramp.xds)
J = make_residence_matrix(3, c(1,1,2,2))
J
```
If $H = \left[10, 100, 20, 200 \right]$ is a column vector describing the size of the strata, then $J\cdot H$ is a vector of length $N_p$ that is the resident population size for all three patches: 

```{r}
J %*% matrix(c(10, 100, 20, 200), 4, 1)
```