---
title: "Trace Functions"
subtitle: "Basic Forcing by Trivial Modules"
editor:
  mode: source
bibliography: ../refs/refs.bib
csl: ../refs/plos.csl
format: html
html-math-method:
  method: mathjax
  url: "https://cdn.jsdelivr.net/npm/mathjax@3/es5/tex-mml-chtml.js"
---

# Overview 

Malaria (and other mosquito-transmitted pathogens) can be understood as *complex adaptive systems* ---non-linear interactions among humans, parasites, mosquitoes, and managers that are forced by weather, hydrology, and other factors.
Dynamical systems have thus been developed to understand malaria epidemiology, malaria transmission dynamics, mosquito ecology, and malaria control. 
These models are naturally modular [@WuSL2023SpatialDynamics]. 
Given the complexity, it is sometimes useful to isolate a part of the system and reduce inputs from the rest of the system to a *trace function.*

We have been using the term *trace function* to describe any function that sets the value of the a dynamical term that has been passed *from* one of the dynamical components. This makes it possible to study sub-models, to study effects of exposure, forcing by malaria control, weather, or biotic factors. 

## Trivial Modules

A trace function sets the value of a **dynamical term.** The *trace functions* live in the **trivial modules** in **`ramp.xds`**. Five terms can be set up as trace functions:

+ $\Lambda(t)$ or `Lambda` --- the emergence rate of adult, female mosquitoes, per patch is passed from the [trivial **L** module](trivialX.qmd).

+ $\eta(t)$ or `eta` --- eggs deposition rate, the number of eggs laid in aquatic habitats passed from the [trivial **MY** module](trivialMY.qmd).

+ $fqZ(t)$ or `fqZ` --- the net rate of infectious biting by the adult mosquito population, per patch is passed from the [trivial **MY** module](trivialMY.qmd).

+ $\kappa(t)$ or `kappa` --- the net infectiousness (NI), the probability that a mosquito will become infected after blood feeding on a human in that patch is passed from the [trivial **X** module](trivialX.qmd).

+ $E(t)$ or `eir` --- the daily entomological inoculation rate (EIR) for a cohort of humans (or the force of infection in a population or a cohort, FoI) is set up by `xds_setup_eir`. The EIR is passed to the **XH** module by a trace function from the EIR object.

The first four terms are passed from trivial modules of the dynamical components. Examples are are documented in their vignettes. Forcing by the **EIR** is handled slightly differently. In @sec-eir: **EIR forcing** [(below)](#sec-eir), we present some fully worked examples that show how to build and use forcing functions. 

## Composed Time Series 

To build in some flexibility in a way that is easy for users to work with, all trace functions are called as **composed time series functions** $x(t)$ that is the product of four elements: 


$$
x(t) = \bar x \times F_S(t) \times F_T(t) \times F_K(t)
$$

where

+ $\bar x$ is a scaling parameter. It is close to the mean value of an unperturbed system. It is the mean over an interval $(t_0, t_1)$ only if $F_K(t)=1$ and $$\int_{t_0}^{t_1} F_S(t)\; F_T(t)\; dt = t_1-t_0$$

+ $F_S(t)>0$ is a **seasonal** pattern function, normalized over the interval such that  $$\int_{t_0}^{t_1} F_S(t) \;dt = t_1 - t_0$$  

+ $F_T(t)>0$ is a **trend** pattern function, normalized over the interval such that $$\int_{t_0}^{t_1} F_T(t) \; dt = t_1 - t_0$$  

+ $F_K(t)>0$ is a **shock** function

## Seasonal Patterns

+ The seasonal pattern function $F_S(t,V)$ is always called `F_season`. 

+ A library of seasonality pattern functions is maintained as part of [`ramp.func`](https://dd-harp.github.io/ramp.func/articles/Seasonality.html)


## Trends
+ The trend pattern function $F_T(t,V)$ is always called `F_trend`. 

+ A library of trend pattern functions is maintained as part of [`ramp.func`](https://dd-harp.github.io/ramp.func/articles/Trends.html)

## Shocks

+ The shock function $F_K(t,V)$ is always called `F_shock`. 

+ A library of shock functions is maintained as part of [`ramp.func`](https://dd-harp.github.io/ramp.func/articles/Shock.html)


# EIR Forcing {#sec-eir}

To understand malaria epidemiology in relation to exposure, we can use [`ramp.xds::xds_setup_eir()`](https://dd-harp.github.io/ramp.xds/reference/xds_setup_eir.html) to construct a trace function. 

::: {.callout-tip title = "Forcing for Cohorts" collapse="true"}
WRITE ME. 
:::

Two ways of building trace functions are: 

+ **User Defined Functions:** Build a trace function and set it up: 

    - Use `xds_setup_eir()`
    
    - Use `change_F_season()`

+ Use `ramp.func::make_function` 

+ Use `ramp.library::change_F_s`

```{r}
library(ramp.xds)
```

## User Defined Functions 

Any function can be configured by a user used for a seasonal pattern, a trend, or a shock. Function has have the form $F(t, V);$ the second variable, $V,$ is part of the flexible and extensible design. 

::: {.callout-tip title = "More about F(t,V)" collapse="true"}
In some cases, users might want to set up an exogenous forcing variable and then call a *trace functions* that uses it, so all the components of the composed time series function have the form $F(t,V).$ In the function calls that evaluate the trace function, the second variable, $V,$ is retrieved by the function `ramp.xds::get_variables` so that trace functions can use any variable stored on the `xds` model object. By default $V$ is an empty list, and `get_variables` returns an empty list
:::

If a user defined function does not need $V,$ the function must define the variable anyway and set its default value to an empty list, like this:

```{r}
F_eg = function(t, V=list()){1 + sin(2*pi*t/365)}
```

The function uses only $t,$ ignoring the second argument (even if it's Santa Claus).

```{r}
t <- c(0:730)
plot(F_eg(t, "Santa Claus"), type = "l", 
     xlab = "Time (Days)", 
     ylab = expression(F[eg]))
```

Having defined `F_eg`, we can use it. In this case, we pass it as the `F_season` argument of `xds_setup_eir`. The function `show_season` plots it (by default, for two full cycles).

```{r}
mod_eg <- xds_setup_eir(F_season = F_eg)
show_season(mod_eg)
```

## `change_F_*` 

`change_F_season` | `change_F_trend`| `change_F_shock`

*** 

If we didn't pass `F_season` to `xds_setup_eir`, the default function is `F_one`, a function that returns the value $1$. 

```{r}
mod_eg <- xds_setup_eir()
show_season(mod_eg)
```

We can modify the `F_season` argument using `change_F_season` 

```{r}
mod_eg <- change_F_season(F_eg, mod_eg)
show_season(mod_eg)
```

::: {.callout-tip title = "`change_F_season`" collapse="true"}
The function `change_F_season` dispatches on `xds_obj$forced_by`, which is set up by the trivial module. 

In this case, the `trivial` **L** module is the default for `xds_setup,` and since no value for `Lname` was passed as an argument, `mod_eg` set up a trivial **L** module. No `Loptions` were passed at setup, so the function `F_season` was originally set to `F_one`.  
:::

Similarly, a user-defined trend or shock could be added using `change_F_trend` or `change_F_shock` and inspected using `show_trend` or `show_shock`  


## `make_function` 

In the function library **`ramp.func`**, [make_function](https://dd-harp.github.io/ramp.func/reference/make_function.html) returns a properly formatted trace function. To use `make_function`, the user passes a function object [`F_obj`](https://dd-harp.github.io/ramp.func/reference/F_obj.html) that was constructed by `makepar_F_methodname` where `methodname` is one of the built-in function families. 

One advantage of these function families is that the functions include shape parameters, and the returned function is normalized (by default) so that it has an average value of $1,$ 

```{r}
library(ramp.func)
par_spline <- makepar_F_spline(tt = c(0:5)*365, yy = c(.1, 1.4, 1, 1.2, 0.5, 1), X=2)
par_sin <- makepar_F_sin(pw=2)
```

Now we make the functions and examine them and their product:

```{r}
t5y <- seq(0, 365*5, by = 5)
F_spline <- make_function(par_spline)
F_sin <- make_function(par_sin)
plot(t5y, F_spline(t5y)*F_sin(t5y), type = "l", xlab = "Time (in Days)", ylab =expression(list(F[sin], F[spline])), lwd=2)
lines(t5y, F_sin(t5y), col = grey(0.5))
lines(t5y, F_spline(t5y), col = grey(0.5))
```

and we can set it up as before. This time, we set the scaling parameter `eir` to have an annual value of 1 infectious bite, per person, per year. 

```{r}
mod_both <- xds_setup_eir(eir=1/365, F_trend = F_spline, F_season=F_sin)
mod_both <- xds_solve(mod_both, times = t5y)
xds_plot_PR(mod_both)
```

## `change_*` 

`change_season` | `change_trend`| `change_shock`

*** 

The `change_*` functions in **`ramp.forcing`** work with `F_obj,` rather than with functions. 

```{r}
library(ramp.forcing)
par_shock <- makepar_F_sharkbite(D=365, L=365)
mod_new <- change_shock(par_shock, mod_both)
show_shock(mod_new)
```

```{r}
mod_new <- xds_solve(mod_new, times = t5y)
xds_plot_PR(mod_both, clrs=grey(0.5))
xds_plot_PR(mod_new, add=T)
```


# I/O

The **`xds`** object can be saved to a file using `saveRDS` and read from a file using `readRDS`. In some cases, the forcing functions `F_season` and `F_trend` and `F_shock` cause the size of the saved file to be very large. 

To save space, [`saveXDS`](file:///Users/smitdave/git/ramp.xds/docs/reference/saveXDS.html) strips the functions before calling `saveRDS`. 

Alternatively, the functions can be set up as function objects `ramp.func`. In this case, the  function `ramp.forcing::readXDS` will read the object using `readRDS` and then rebuild the functions using `ramp.forcing::rebuild_forcing_functions().`



