---
title: "The `ramp.xds` Pipeline"
subtitle: "An interactive tutorial: setup → solve → visualize"
format: html
server: shiny
---

------------------------------------------------------------------------

*This tutorial walks through the three-step **`ramp.xds`** pipeline interactively. Adjust parameters using the controls below and watch the model update in real time.*

------------------------------------------------------------------------

```{r}
#| context: setup
#| include: false
library(ramp.xds)
```

## Overview

Every **`ramp.xds`** workflow follows the same three steps:

1.  **Setup** — `xds_setup()` configures a model object by choosing dynamical modules and setting parameters.
2.  **Solve** — `xds_solve()` integrates the system of differential equations and stores outputs.
3.  **Visualize** — plotting functions like `xds_plot_PR()` display results from the solved model.

------------------------------------------------------------------------

## Step 1 — Setup

`xds_setup()` returns an **`xds` model object**: a fully configured model ready to solve. Every argument has a default, so `xds_setup()` with no arguments immediately returns a working model. The three **dynamic components** are each chosen by name:

- **XH** — human demography, infection dynamics & immunity: `Xname` (default `"SIS"`)
- **MY** — adult mosquito ecology & infection dynamics: `MYname` (default `"SI"`)
- **L** — aquatic mosquito ecology: `Lname` (default `"trivial"`)

This tutorial uses the classic **Ross–Macdonald** formulation: `Xname = "SIS"` paired with `MYname = "macdonald"`.

Adjust parameters below. The live `xds_setup()` call reflects your choices.

```{r}
#| echo: false
fluidRow(
  column(4,
    wellPanel(
      h4(HTML("<b>Human Infection (XH)</b> &mdash; <code>SIS</code>")),
      sliderInput("b",
                  HTML("<em>b</em> &mdash; probability of infection per infectious bite:"),
                  min = 0.1, max = 0.9, value = 0.55, step = 0.05),
      sliderInput("r_inv",
                  HTML("1/<em>r</em> &mdash; mean infection duration (days):"),
                  min = 30, max = 365, value = 180, step = 10),
      sliderInput("cc",
                  HTML("<em>c</em> &mdash; human infectiousness to mosquitoes:"),
                  min = 0.05, max = 0.50, value = 0.15, step = 0.05),
      sliderInput("HPop",
                  HTML("<em>H</em> &mdash; human population size:"),
                  min = 100, max = 5000, value = 1000, step = 100)
    )
  ),
  column(4,
    wellPanel(
      h4(HTML("<b>Adult Mosquitoes (MY)</b> &mdash; <code>macdonald</code>")),
      sliderInput("f",
                  HTML("<em>f</em> &mdash; blood feeding rate (per day):"),
                  min = 0.1, max = 1.0, value = 0.3, step = 0.05),
      sliderInput("g_inv",
                  HTML("1/<em>g</em> &mdash; mean mosquito lifespan (days):"),
                  min = 3, max = 30, value = 12, step = 1),
      sliderInput("eip",
                  HTML("EIP &mdash; extrinsic incubation period (days):"),
                  min = 5, max = 25, value = 12, step = 1)
    )
  ),
  column(4,
    wellPanel(
      h4(HTML("<b>Aquatic Mosquitoes (L)</b> &mdash; <code>trivial</code>")),
      sliderInput("Lambda",
                  HTML("&Lambda; &mdash; adult mosquito emergence rate (per day):"),
                  min = 100, max = 5000, value = 1000, step = 100),
      hr(),
      h4(HTML("<b>Solver</b>")),
      sliderInput("Tmax",
                  HTML("<em>T</em><sub>max</sub> &mdash; simulation duration (days):"),
                  min = 60, max = 3650, value = 365, step = 30)
    )
  )
)
```

**Generated `xds_setup()` call:**

```{r}
#| echo: false
verbatimTextOutput("setup_code")
```

------------------------------------------------------------------------

## Step 2 — Solve

`xds_solve()` integrates the dynamical system using `deSolve` and attaches outputs to the model object. The call is straightforward:

```{r}
#| echo: false
verbatimTextOutput("solve_code")
```

The output stores time series for all state variables and computed terms (EIR, force of infection, kappa, etc.) in `model$outputs$orbits`.

------------------------------------------------------------------------

## Step 3 — Visualize

After solving, **`ramp.xds`** plotting functions display each component of the system. Select a tab to explore the outputs.

```{r}
#| echo: false
tabsetPanel(id = "plots",
  tabPanel("Prevalence (PR)",
    br(),
    plotOutput("pr_plot", height = "320px"),
    p(HTML("<small>Parasite rate (PR) &mdash; the fraction of the human population
           that is infected (<em>I/H</em>) &mdash; over time. Plotted using
           <code>xds_plot_PR()</code>.</small>"))
  ),
  tabPanel("EIR",
    br(),
    plotOutput("eir_plot", height = "320px"),
    p(HTML("<small>Entomological Inoculation Rate (EIR) &mdash; the number of
           infectious bites per person per year &mdash; over time. Plotted using
           <code>xds_plot_EIR()</code>.</small>"))
  ),
  tabPanel("Mosquito Density (M)",
    br(),
    plotOutput("M_plot", height = "320px"),
    p(HTML("<small>Total adult mosquito density over time. Plotted using
           <code>xds_plot_M()</code>.</small>"))
  ),
  tabPanel("Infected Mosquitoes (Y)",
    br(),
    plotOutput("Y_plot", height = "320px"),
    p(HTML("<small>Density of mosquitoes that have been infected but have
           not yet completed the extrinsic incubation period. Plotted using
           <code>xds_plot_Y()</code>.</small>"))
  ),
  tabPanel("Infective Mosquitoes (Z)",
    br(),
    plotOutput("Z_plot", height = "320px"),
    p(HTML("<small>Density of infective mosquitoes &mdash; those carrying
           sporozoites and capable of transmitting malaria. Plotted using
           <code>xds_plot_Z()</code>.</small>"))
  )
)
```

------------------------------------------------------------------------

```{r}
#| context: server

# --- Reactive: build the model -----------------------------------------------
mod <- reactive({
  r <- 1 / input$r_inv
  g <- 1 / input$g_inv
  xds_setup(
    Xname     = "SIS",
    MYname    = "macdonald",
    Lname     = "trivial",
    HPop      = input$HPop,
    XHoptions = list(b = input$b, r = r, c = input$cc),
    MYoptions = list(f = input$f, g = g, eip = input$eip),
    Loptions  = list(Lambda = input$Lambda)
  )
})

# --- Reactive: solve the model -----------------------------------------------
solved_mod <- reactive({
  xds_solve(mod(), Tmax = input$Tmax, dt = 1)
})

# --- Show the generated xds_setup() call -------------------------------------
output$setup_code <- renderText({
  r_val <- signif(1 / input$r_inv, 4)
  g_val <- signif(1 / input$g_inv, 4)
  paste0(
    'model <- xds_setup(\n',
    '  Xname     = "SIS",\n',
    '  MYname    = "macdonald",\n',
    '  Lname     = "trivial",\n',
    '  HPop      = ', input$HPop,    ',\n',
    '  XHoptions = list(b = ', input$b,   ', r = ', r_val,  ', c = ', input$cc,  '),\n',
    '  MYoptions = list(f = ', input$f,   ', g = ', g_val,  ', eip = ', input$eip, '),\n',
    '  Loptions  = list(Lambda = ', input$Lambda, ')\n',
    ')'
  )
})

# --- Show the generated xds_solve() call -------------------------------------
output$solve_code <- renderText({
  paste0('model <- xds_solve(model, Tmax = ', input$Tmax, ', dt = 1)')
})

# --- Plots -------------------------------------------------------------------
output$pr_plot <- renderPlot({
  xds_plot_PR(solved_mod(), y01 = TRUE)
})

output$eir_plot <- renderPlot({
  xds_plot_EIR(solved_mod())
})

output$M_plot <- renderPlot({
  xds_plot_M(solved_mod())
})

output$Y_plot <- renderPlot({
  xds_plot_Y(solved_mod())
})

output$Z_plot <- renderPlot({
  xds_plot_Z(solved_mod())
})
```
